Probabilistic bioregion models suggest seabird assemblages are linked to the major boundary current off eastern Australia

### 2023-07-11 docs/ms-draft.Rmd
small adjustments on fig numbering and I've sent this version off to coauthors

Given east Australia's importance for seabirds and the recent attention on how climate change in the EAC can affect marine biota [@poloczanska2007], an assessment of the spatial distribution of seabirds at large scales is timely. Seabirds can serve as sentinels for tracking ecological changes in the EAC in many ways [e.g. @sydeman2012]. Surprisingly, though, seabird distribution is under-represented in climate change studies, despite being highlighted as a priority topic [@grémillet2009; @sydeman2012]. In this study, we aim to identify assemblages of seabird species probabilistically and relate them to environmental conditions, by using at-sea surveys data and a state-of-the-art modelling technique. We used the mixture-model Region of Common Profiles [RCP; @foster2013], regarded as one of the best available tools for describing bioregions [@hill2020; @woolley2020]. RCP is a model-based approach, and has a range of advantages over traditional algorithm-based statistics [@warton2015; @hill2020]. East Australia supports a high diversity of seabirds and offers wide environmental gradients, from the tropics to the subantarctic. Therefore, we expect to find different assemblages relating to different environmental conditions. In addition, we run RCP models for presence-absence and abundance as input data, to test if the different numeric resolution would result in similar assemblage patterns. Furthermore, we provide a repeatable framework on which future studies can build to assess how changes in the EAC will influence the distribution of seabird assemblages.

@@

Spatial and temporal persistence of features such as eddy kinetic energy (eke) and thermal fronts (sea surface temperature gradient; sst_grad) may influence the movement and distribution of seabirds by offering a higher likelihood of finding food [@weimerskirch2007; @scales2014a]. Therefore, before extracting environmental data for each seabird record, 'eke' and 'sst_grad' environmental layers were calculated. For both, 'eke' and 'sst_grad', we used their original cell spatial resolution (**Table 1**) and calculations were done layer-by-layer to match seabird data on the finest scale possible. Eddy kinetic energy was calculated as $0.5 (u'² + v'²)$, where $u' = u - \bar{u}$ and $v' = v - \bar{v}$ are the surface zonal (latitudinal) and meridian (longitudinal) geostrophic currents, respectively, using a rolling mean of 90 days as $\bar{u}$ and $\bar{v}$ (**ROB: Ref here?**). The spatial 'sst_grad' for each cell was calculated by averaging the maximum value for each neighbouring cell (average neighbourhood maximum), from north-to-south (NS) and west-to-east (WE); then, the magnitude of the gradient was defined as $NS² + WE²$ [@burrows2011]. In addition, we calculated climatic seasonal average layers for 'eke' and 'sst_grad' based on 10-years of data (2011--2021) as a proxy for persistent oceanographic features, as well as 'eke' standard deviation over the same period (named with the prefix 'clim\_' in **Table 1**). Seabed slope was calculated using 'raster::terrain()', and distance from the coast was calculated through 'sf::st_distance()' using Australia and Tasmania as the mainland (i.e. ignoring smaller islands). Then, for each seabird record, the most proximate value for each environmental layer (original or calculated) was extracted using built-in functions from 'raster', 'terra', and 'sf' packages.

@@

For all models two RCPs was the optimal number of groups, except for presence-absence-autumn, which indicated three RCPs as the best fit (**Fig S3**). The best fitting models generally had 1 or 2 covariates selected, although the abundance-autumn model had five covariates (**Table 2**). The best models showed a good fit with no apparent deviations from the assumptions (**Fig. S4**). All the best models selected 'sst' or 'sss', variables essential for characterising water masses. The probability of belonging to one or another RCP varied its 'sst' threshold between 18--21°C, or 'sss' 35.1 PSU for summer (**Fig. 4**, **Fig. S6**), similar to EAC's temperature and salinity thresholds. Furthermore, most of the best models selected 'bat', suggesting a gradient between 'coastal' and 'oceanic' taxa (**Fig. S6**). Interestingly, proxies for persistent oceanographic features that may aggregate food resources (the 'climatic' variables) were selected only in autumn and spring abundance models.

@@

```{r fig4-partial-plots, fig.cap = "**FIGURE 4.** Partial plots for the retained covariates in the spring (a) presence-absence and (b) abundance best models. Results for other seasons and models can be found in the Supplementary Material (Fig. S6). Refer to Table 1 for the environmental data acronyms.", fig.align = "center", echo = FALSE}

@@

Grid point-predictions are the assignment of the RCP with the highest probability value for a given cell. When predicting RCP onto space, patterns of point-predictions suggest a biogeographic change around \~34°S, with both presence-absence and abundance models pointing to similar patterns (**Fig. 2**). These patterns are further reinforced when looking at the mean probability along with the lower and upper confidence intervals of predictions (**Fig. 3**, **Fig. S5**), showing that around \~34°S the 'northern' RCP tends to have higher probabilities in coastal regions, whereas the 'southern' RCP spreads northward offshore. The presence-absence-autumn model (three RCPs) has a similar broad-scale pattern, although the southern RCP is further south, and a strong 'Tasmania/Bass Straight' assemblage seems to appear (**Fig. 2**, **Fig. S5**).

```{r fig2-point-pred, fig.cap = "**FIGURE 2.** Spatialized point predictions (hard-classes) for seabird assemblages (Region of Common Profile; RCP), by season, based on presence-absence (a) and abundance (b) models, off eastern Australia. Each grid was assigned to the RCP group with highest membership probability.", fig.align = "center", echo = FALSE}

@@

```{r fig3-prob-maps, fig.cap = "**FIGURE 3.** Predicted probability membership of for each seabird assemblage (Region of Common Profiles; RCP) and grid, off eastern Australia, from spring presence-absence (a) and abundance (b) models. The central column, 'mean', corresponds to the point prediction and Bayesian bootstraped, lower and upper confidence intervals (CI), on its sides. Results for other seasons and models can be found in the Supplementary Material (Fig. S5).", fig.align = "center", echo = FALSE}

@@

```{r fig5-spp-profiles, fig.cap = "**FIGURE 5.** Species profiles of seabird assemblages (Region of Common Profile; RCP) off eastern Australia in spring. Values are the average and 95% confidence intervals of probability of occurrence (a) and predicted mean abundance (b) for each species, based on 1000 Bayesian bootstraps. Values in (b) were log10-transformed to accommodate the high variation between species. Results for other seasons and models can be found in the Supplementary Material.", fig.align = "center", echo = FALSE}

@@

The seasonal models also captured the influence of migratory species, particularly in the southern assemblage. For instance, summer and autumn models reflected the high abundance of the short-tailed shearwater *Ardenna tenuirostris* before they depart on their trans-equatorial migration [@carey2014]; the presence of the Australasian gannet was picked up by autumn and winter models [@ismar2011]; and the white-headed petrel *Pterodroma lessonii* movements during chick-rearing and pre-laying exodus [@taylor2021] were detected in autumn and spring models, respectively. Given the geographic location of the southern assemblage, however, we believe subantarctic taxa were not well represented in our models due to their low number of occurrences (**Fig. S1 - FO/nOCC/FN**; hence they were removed before modelling). More intensive surveys may reveal the seasonal northward pulse of these taxa into the southern assemblage (e.g. albatrosses *Diomedea* spp. and *Phoebetria* spp.), given that more effort translates as higher likelihood of recording these species. In turn, including these taxa in the models could result in a more stable 'three assemblage pattern' (as revealed by the presence-absence-autumn model) in all but the summer season, when the majority of Antarctic species are breeding **[Harrison 1983 - is there a better ref here, seabirders?]**.

>>>>

Given east Australia's importance for seabirds and the recent attention on how climate change in the EAC can affect marine biota [@poloczanska2007], an assessment of the spatial distribution of seabirds at large scales is timely. Seabirds can serve as sentinels for tracking ecological changes in the EAC in many ways [e.g. @sydeman2012]. Surprisingly, though, seabird distribution is under-represented in climate change studies, despite being highlighted as a priority topic [@grémillet2009; @sydeman2012]. In this study, we aim to identify assemblages of seabird species probabilistically and relate them to environmental conditions, by using at-sea surveys data and a state-of-the-art modelling technique. We used the mixture model Region of Common Profiles [RCP; @foster2013], regarded as one of the best available tools for describing bioregions [@hill2020; @woolley2020]. RCP is a model-based approach, and has a range of advantages over traditional algorithm-based statistics [@warton2015; @hill2020]. East Australia supports a high diversity of seabirds and offers wide environmental gradients, from the tropics to the subantarctic. Therefore, we expect to find different assemblages relating to different environmental conditions. In addition, we run RCP models for presence-absence and abundance as input data, to test if the different numeric resolution would result in similar assemblage patterns. Furthermore, we provide a repeatable framework on which future studies can build to assess how changes in the EAC will influence the distribution of seabird assemblages.

@@

Spatial and temporal persistence of features such as eddy kinetic energy (eke) and thermal fronts (sea surface temperature gradient; 'sst_grad') may influence the movement and distribution of seabirds by offering a higher likelihood of finding food [@weimerskirch2007; @scales2014a]. Therefore, before extracting environmental data for each seabird record, 'eke' and 'sst_grad' environmental layers were calculated. For both, 'eke' and 'sst_grad', we used their original cell spatial resolution (**Table 1**) and calculations were done layer-by-layer to match seabird data on the finest scale possible. Eddy kinetic energy was calculated as $0.5 (u'² + v'²)$, where $u' = u - \bar{u}$ and $v' = v - \bar{v}$ are the surface zonal (latitudinal) and meridian (longitudinal) geostrophic currents, respectively, using a rolling mean of 90 days as $\bar{u}$ and $\bar{v}$ (**ROB: Ref here?**). The spatial 'sst_grad' for each cell was calculated by averaging the maximum value for each neighbouring cell (average neighbourhood maximum), from north-to-south (NS) and west-to-east (WE); then, the magnitude of the gradient was defined as $NS² + WE²$ [@burrows2011]. In addition, we calculated climatic seasonal average layers for 'eke' and 'sst_grad' based on 10-years of data (2011--2021) as a proxy for persistent oceanographic features, as well as 'eke' standard deviation over the same period (named with the prefix 'clim\_' in **Table 1**). Seabed slope was calculated using 'raster::terrain()', and distance from the coast was calculated through 'sf::st_distance()' using Australia and Tasmania as the mainland (i.e. ignoring smaller islands). Then, for each seabird record, the most proximate value for each environmental layer (original or calculated) was extracted using built-in functions from 'raster', 'terra', and 'sf' packages.

@@

For all models two RCPs was the optimal number of groups, except for presence-absence-autumn, which indicated three RCPs as the best fit (**Fig S3**). The best fitting models generally had 1 or 2 covariates selected, although the abundance-autumn model had five covariates (**Table 2**). The best models showed a good fit with no apparent deviations from the assumptions (**Fig. S4**). All the best models selected 'sst' or 'sss', variables essential for characterising water masses. The probability of belonging to one or another RCP varied its 'sst' threshold between 18--21°C, or 'sss' 35.1 PSU for summer (**Fig. 2**, **Fig. S5**), similar to EAC's temperature and salinity thresholds. Furthermore, most of the best models selected 'bat', suggesting a gradient between 'coastal' and 'oceanic' taxa (**Fig. S5**). Interestingly, proxies for persistent oceanographic features that may aggregate food resources (the 'climatic' variables) were selected only in autumn and spring abundance models.

@@

```{r fig4-partial-plots, fig.cap = "**FIGURE 2.** Partial plots for the retained covariates in the spring (a) presence-absence and (b) abundance best models. The plot shows the probability of belonging to a Region of Common Profiles (RCP) against the environmental value. Results for other seasons and models can be found in the Supplementary Material (Fig. S5). Refer to Table 1 for the environmental data acronyms.", fig.align = "center", echo = FALSE}

@@

Grid point-predictions are the assignment of the RCP with the highest probability value for a given cell. When predicting RCP onto space, patterns of point-predictions suggest a biogeographic change around \~34°S, with both presence-absence and abundance models pointing to similar patterns (**Fig. 3**). These patterns are further reinforced when looking at the mean probability along with the lower and upper confidence intervals of predictions (**Fig. 4**, **Fig. S6**), showing that around \~34°S the 'northern' RCP tends to have higher probabilities in coastal regions, whereas the 'southern' RCP spreads northward offshore. The presence-absence-autumn model (three RCPs) has a similar broad-scale pattern, although the southern RCP is further south, and a strong 'Tasmania/Bass Straight' assemblage seems to appear (**Fig. 3**, **Fig. S6**).

```{r fig2-point-pred, fig.cap = "**FIGURE 3.** Spatialized point predictions (hard-classes) for seabird assemblages (Region of Common Profile; RCP), by season, based on presence-absence (a) and abundance (b) models, off eastern Australia. Each grid was assigned to the RCP group with highest membership probability.", fig.align = "center", echo = FALSE}

@@

```{r fig3-prob-maps, fig.cap = "**FIGURE 4.** Predicted probability membership of for each seabird assemblage (Region of Common Profiles; RCP) and grid, off eastern Australia, from spring presence-absence (a) and abundance (b) models. The central column, 'mean', corresponds to the point prediction and Bayesian bootstraped, lower and upper confidence intervals (CI), on its sides. Results for other seasons and models can be found in the Supplementary Material (Fig. S6).", fig.align = "center", echo = FALSE}

@@

```{r fig5-spp-profiles, fig.cap = "**FIGURE 5.** Species profiles of seabird assemblages (Region of Common Profile; RCP) off eastern Australia in spring. Values are the average and 95% confidence intervals of probability of occurrence (a) and predicted mean abundance (b) for each species, based on 1000 Bayesian bootstraps. Values in (b) were log10-transformed to accommodate the high variation between species. Results for other seasons and models can be found in the Supplementary Material (Fig. S7).", fig.align = "center", echo = FALSE}

@@

The seasonal models also captured the influence of migratory species, particularly in the southern assemblage. For instance, summer and autumn models reflected the high abundance of the short-tailed shearwater *Ardenna tenuirostris* before they depart on their trans-equatorial migration [@carey2014]; the presence of the Australasian gannet was picked up by autumn and winter models [@ismar2011]; and the white-headed petrel *Pterodroma lessonii* movements during chick-rearing and pre-laying exodus [@taylor2021] were detected in autumn and spring models, respectively. Given the geographic location of the southern assemblage, however, we believe subantarctic taxa were not well represented in our models due to their low number of occurrences (**Fig. S1 - FO/nOCC/FN**; hence they were removed before modelling). More intensive surveys may reveal the seasonal northward pulse of these taxa into the southern assemblage (e.g. albatrosses *Diomedea* spp. and *Phoebetria* spp.), given that more effort translates as higher likelihood of recording these species. In turn, including these taxa in the models could result in a more stable 'three assemblage pattern' (as revealed by the autumn presence-absence model) in all but the summer season, when the majority of Antarctic species are breeding **[Harrison 1983 - is there a better ref here, seabirders?]**.
